Consistency and Fault Tolerance Considerations for the Next Iteration of the DOE Fast Forward Storage and IO Project

### 2014-03-21 paper.tex
added comments from Quincey and expanded acknowledgements.

While these and other their contributions are valuable, similar efforts made
elsewhere may offer attractive alternatives or differing semantics that could
yield a more feature rich environment with little to no additional overhead.
For example, the Doubly Distributed Transactions (\DDTns) protocol offers an
alternative approach for incorporating transactional semantics into the data
path. The PreDatA project examined how to get the best throughput for data
operators can offer some insights for refining the Burst Buffer concept.

@@

The core idea for IOD is to provide a way to manage the IO load that is
separate from the compute nodes and the storage array. Communication intensive
activities, such as two-phase, data sieving IO's data rearrangement process,
can be moved to the IOD layer offloading the communication load from the
compute nodes. IOD has three main purposes. First, if the optional burst buffer
is available, it works as a fast cache absorbing write operations for the
slower trickle out to the central storage array. It can also be used to
retrieve objects from the central storage array for more efficient read
operations and offer data filtering to make client reads more efficient.
Second, it offers the transaction mechanism for controlling data set visibility
and to manage faults that would prevent a data set from being used. Third, data
processing operations can be placed in the IOD. These operations are intended
to offer data rearrangement, filtering, and similar operations prior to data
reaching the central storage array.

@@

Burst buffers being optional is a high level goal, but not one considered at a
detailed level with the design. While this forces some serious thought on how
to adapt the functionality proposed for IOD to work directly on DAOS will have
to be considered. With the additional desire to support using compute node
resources for these operations, serious work will be required to make a fully
functional end-to-end IOD layer implementation for a production system.

Another concern that is acknowledged, but no thought as been applied to is the
requirement that a single IOD process be the master for any operation. Should
the number of concurrent applications exceed the available nodes, sharing an
IOD process will be required. Since partitioning of the IOD processes for
exclusive use by particular applications is the assumed operating mode, should
insufficient IOD resources be available, either a job could be delayed or IOD
resources could be reallocated from a different process could be redeployed for
use by the new job. These sorts of considerations still need to be made for a
full production system.

@@

\subsection{IOD Transactions}
To understand how transactions are used in the IOD layer, some terminology and
concepts must be explained first. At the coarsest grain level is a container.
Each container serves to host a transaction and contains a collection of
objects. Conceptually, containers corresponds to a file in a traditional file
system. The objects in each container represent different data within a file.
The three initially defined object types are key-value store, array shard,
and blob. The easiest way to understand these types is to evaluate these from
the perspective of an HDF-5 file, the initial user interface layer. The
key-value store represents a collection of attributes. The array shard
represents part of a potentially multi-dimensional array. It is referred to as
a shard because it is likely a small piece of a globally defined array. The
blob represents a byte stream of arbitrary contents.  The fundamental
difference between an array shard and a blob is that the array shard has
metadata identifying its portion within the global, logical space while the
blob is simply a 1-dimensional array of bytes that is not shared across IOD
nodes.  Should an operation be deployed into IOD to manipulate an array within
a container, it would operate on the array shards rather than on blobs. Given
this context, the transactions come in two forms.

@@

The second form is called multi-leader and has the IOD layer manage the
transactions. In this case, when the transaction is created, a count of clients
is provided to the IOD layer. As clients write to the container, the reference
count is reduced. Once the count reaches 0, the transaction is automatically
committed.

@@

A new term, ``Primary'', is used to define the process(es) that contact the IOD
layer directly to open a container. Unfortunately, this term is never used nor
defined within any of the published documents. There are some very positive and
negative performance implications of this approach. On the positive side, like
PVFS, it is possible for a single process to open a container and share that
handle with other processes reducing the load on the IOD layer. The key idea is
that the single leader vs. multi-leader transaction mode is automatically
defined based on the number of Primary processes that open a container. This
greatly simplifies any additional client-side load in using the different
styles of transactions. Also on the positive side, it offers the ability to
have a subset of processes that are considered transaction participants
affording reduced load on the IOD processes and more localized transaction
management on the client side. On the negative side, for a fully IOD managed
transaction, ALL processes must open the container forcing a potentially
immense number of simultaneous calls to the IOD layer to open a container. This
sort of onslaught is what is dealt with on metadata servers today with mass
file opens. The reality is that this ends up having the same sort of
performance implications as distributed denial of service attacks on the IOD
processes.

Ultimately, with the passive detection of faults for Primaries, the transaction
mechanism can work very well. A mostly unstated restriction that will be
relaxed in the next phase is that every sequential transaction on a container
is considered dependent on the earlier transaction. Should one output be
delayed and the subsequent five succeed, when the delayed process finally
fails, all six transactions are rolled back. The thought of using this
mechanism to store subsequent checkpoint outputs in the same container to both
save space, but not care if one fails, cannot work in the current form. This
has been acknowledged and is planned on being relaxed in phase two.

@@

\subsection{Metadata Management}
Eliminating metadata management as a special case and instead treating it just
as data is a central design goal of the Fast Forward project. Unfortunately,
it is not completely achieved.

@@

With a client-side only approach, there is a serious risk of the metadata
service and the object store getting out of sync.  While having a metadata-less
object storage service is desirable, the different semantics from traditional
file systems requires some considerations. In this case, should these services
get out of sync, three particular risks introduced.  First, a client could
create an entry in the metadata service that does not correspond to any objects
in the object store. Second, a client could create objects that have no
associated metadata entries. Third, updates to the metadata or object store
service should be an atomic operation, but due to a lack of coordination, a
window where the system is inconsistent appears.  Ultimately, the consistency
semantics required must be determined. If a metadata service is required and it
must be in sync with the object storage service, then additional work must be
performed. In traditional file systems, the metadata and object storage updates
are atomic. With decoupling metadata from object storage, should this atomicity
still be desired, it requires both the ability for the services to participate
in a task that is part of a larger atomic operation and a higher-level
mechanism to manage the atomic operation. Overall, while additional work is
required to maintain a client-side only metadata service, it eliminates any
potential bottlenecks related to updating metadata related to the object
storage. The burden of tracking striping and other metadata that has
traditionally been part of the metadata associated with the file system will
have to be maintained by the object storage service. The lack of a centralized,
serialized bottleneck to store that information improves concurrency.

@@

In the multi-leader model, using a count of client connections to determine if
transaction is complete is problematic.  First, the aggregation of completion
messages are based on the local aggregation point reporting to the transaction
leader.  There is no ability to detect or recover from a failure of this IOD
process that would report to the transaction leader. The current assumption
that the IOD layer will be an MPI program is not workable until MPI can fully
support fault detection and recovery including removing and adding processes.
Second, because a simple count of participating processes is sent with the
beginning of the transaction, each process is strictly limited to a single
output operation.  This precludes a process writing multiple array shards, for
example. To put this in concrete terms, it means that a file could only contain
a single, globally distributed variable (at the root level of HDF5) OR an
attribute per process. If this count actually represents a begin/end
transaction pair instead, it suffers from a lack of accountability for missing
object (shard) writes. All it would indicate is that a process started and
ended a transaction connection. There is no validation that a process has done
everything required of it.

@@

\DDT has addressed theses issues in a couple of ways. First, the
sub-coordinators each have a list of processes from which they expect messages.
Should a message be missed, it is noticed and corrective action can be taken.
Second, \DDT has the concept of sub-transactions. The messaging requirements
are illustrated in Figure~\ref{fig:optimized-protocol}.  Sub-transactions
represent finer grained operations than the entire output, \DDT can manage
multiple writes per client by using a sub-transaction to represent the output
for any item to the file (container). Because of how the sub-transactions are
managed, the singleton sub-transactions, ones in which only a single process
participates, must be declared before the transaction begins so that its
existence can be broadcast as part of the begin transaction message. This
ensures there is global knowledge that the sub-transaction is expected.  That
way if the coordinator (transaction leader) fails, which ever process takes
over that role knows to expect a completion message for that sub-transaction or
the overall transaction cannot complete. Global sub-transactions can be defined
at any time since they are a global, synchronized operation broadcasting their
existence. While this additional layer does introduce messaging, the overhead
is quite small.

@@

At a broader level, there are some concerns that were partially clarified
through conversations with the team.  Consider a shared file system across an
HPC data center. The current design maintains the metadata in its own
container. Since copying data from the IOD layer to the DAOS layer requires an
explicit persist call, how and when synchronizing the metadata across the 
layers and potentially across machines occurs is undefined. By delaying the
synchronization until an explicit persist is called will reduce the update
frequency, it does delay the visibility of data. Ideally, the metadata object
would need to be automatically persisted every time a container transaction is
persisted to the DAOS layer.

@@

\section{Acknowledgements}
\includegraphics[scale=0.07]{logos/doe_logo}
\includegraphics[scale=0.30]{logos/snl_logo}
\includegraphics[scale=0.35]{logos/nnsa_logo}
Sandia National Laboratories is a multi-program laboratory managed and operated
by Sandia Corporation, a wholly owned subsidiary of Lockheed Martin
Corporation, for the U.S. Department of Energy's National Nuclear Security
Administration under contract DE-AC04-94AL85000.

>>>>

While these and their other contributions are valuable, similar efforts made
elsewhere may offer attractive alternatives or differing semantics that could
yield a more feature rich environment with little to no additional overhead.
For example, the Doubly Distributed Transactions (\DDTns) protocol offers an
alternative approach for incorporating transactional semantics into the data
path. The PreDatA project examined how to get the best throughput for data
operators can offer some insights for refining the Burst Buffer concept.

@@

The core idea for IOD is to provide a way to manage the IO load that is
separate from the compute nodes and the storage array. Communication intensive
activities, such as data rearrangement, can be moved to the IOD layer
offloading the communication load from the compute nodes. IOD has three main
purposes. First, if the optional burst buffer is available, it works as a fast
cache absorbing write operations for the slower trickle out to the central
storage array. It can also be used to retrieve objects from the central storage
array for more efficient read operations and offers data filtering to make
client reads more efficient.  Second, it offers the transaction mechanism for
controlling data set visibility and to manage faults that would prevent a data
set from being used. Third, data processing operations can be placed in the
IOD. These operations are intended to offer data rearrangement, filtering, and
similar operations prior to data reaching the central storage array.

@@

Burst buffers being optional is a high level goal, but not one considered at a
detailed level with the design. For example, if there is no burst buffer, all
of the advanced functionality proposed for the IOD layer would have to work
against the DAOS layer instead. For example, function shipping assumes it will
operate on fast, local data within the IOD layer rather than against the
globally shared DAOS layer. With the additional desire to support using compute
node resources for these operations, serious work will be required to make a
fully functional end-to-end IOD layer implementation for a production system.

Another concern that is acknowledged, but no thought has been applied to is the
requirement that a single IOD process be the master for any operation. Should
the number of concurrent applications exceed the available nodes, sharing an
IOD process will be required. Since partitioning of the IOD processes for
exclusive use by particular applications is the assumed operating mode, should
insufficient IOD resources be available, either a job could be delayed or IOD
resources could be reallocated from a different process could be redeployed for
use by the new job. These sorts of considerations still need to be made for a
full production system.

@@

\subsection{IOD Transactions}
To understand how transactions are used in the IOD layer, some terminology and
concepts must be explained first. At the coarsest grain level is a container.
Each container serves to host a transaction and contains a collection of
objects. Conceptually, containers corresponds to a file in a traditional file
system. The objects in each container represent different data within a file.
The three initially defined object types are key-value store, array, and blob.
The easiest way to understand these types is to evaluate these from the
perspective of an HDF-5 file, the initial user interface layer. The key-value
store represents a collection of attributes or groups. The array represents
part of a potentially multi-dimensional array. It is referred to as
a shard because it is likely a small piece of a globally defined array. The
blob represents a byte stream of arbitrary contents.  The fundamental
difference between an array and a blob is that the array has metadata
identifying its portion within the global, logical space while the blob is
simply a 1-dimensional array of bytes that is not shared across IOD nodes.
Should an operation be deployed into IOD to manipulate an array within a
container, it would operate on the array rather than on blobs. Given this
context, the transactions come in two forms.

@@

The second form is called multi-leader and has the IOD layer manage the
transactions. In this case, when the transaction is created, a count of clients
is provided to the IOD layer. As clients commit their changes to the container,
the reference count is reduced. Once the count reaches 0, the transaction is
automatically committed.

@@

The team uses the term ``Primary'' to define the process(es) that contact the
IOD layer directly to open a container. Unfortunately, this term is never used
nor defined within any of the published documents. There are some very positive
and negative performance implications of this approach. On the positive side,
like PVFS, it is possible for a single process to open a container and share
that handle with other processes reducing the load on the IOD layer. The key
idea is that the single leader vs. multi-leader transaction mode is
automatically defined based on the number of Primary processes that start a
transaction on a container. This greatly simplifies any additional client-side
load in using the different styles of transactions. Also on the positive side,
it offers the ability to have a subset of processes that are considered
transaction participants affording reduced load on the IOD processes and more
localized transaction management on the client side. On the negative side, for
a fully IOD managed transaction, ALL processes must start the transaction on
the container forcing a potentially immense number of simultaneous calls to the
IOD layer to open a container. This sort of onslaught is what is dealt with on
metadata servers today with mass file opens. The reality is that this ends up
having the same sort of performance implications as distributed denial of
service attacks on the IOD processes.

Ultimately, with the passive detection of faults for Primaries, the transaction
mechanism can work very well. A mostly unstated restriction that is being
relaxed is that every sequential transaction on a container is considered
dependent on the earlier transaction. Should one output be delayed and the
subsequent five succeed, when the delayed process finally fails, all six
transactions are rolled back. The thought of using this mechanism to store
subsequent checkpoint outputs in the same container to both save space, but not
care if one fails, cannot work in the current form. This has been acknowledged
and is planned on being relaxed in phase two.

@@

\subsection{Metadata Management}
Eliminating metadata management as a special case and instead treating it just
as data is a central design goal of the Fast Forward project. This is a hybrid
approach to metadata management that is half-way between providing no inherent
metadata support and having a fully integrated, but separate metadata
manaegement system.

@@

With a client-side only approach, there is a serious risk of the metadata
service and the object store getting out of sync.  While having a metadata-less
object storage service is desirable, the different semantics from traditional
file systems requires some considerations. In this case, should these services
get out of sync, three particular risks are introduced.  First, a client could
create an entry in the metadata service that does not correspond to any objects
in the object store. Second, a client could create objects that have no
associated metadata entries. Third, updates to the metadata or object store
service should be an atomic operation, but due to a lack of coordination, a
window where the system is inconsistent appears.

Ultimately, the consistency semantics required must be determined. If a
metadata service is required and it must be in sync with the object storage
service, then additional work must be performed. In traditional file systems,
the metadata and object storage updates are atomic. With decoupling metadata
from object storage, should this atomicity still be desired, it requires both
the ability for the services to participate in a task that is part of a larger
atomic operation and a higher-level mechanism to manage the atomic operation.

Overall, while additional work is required to maintain a client-side only
metadata service, it eliminates any potential bottlenecks related to updating
metadata related to the object storage. The burden of tracking striping and
other metadata that has traditionally been part of the metadata associated with
the file system will have to be maintained by the object storage service. The
lack of a centralized, serialized bottleneck to store that information improves
concurrency.

@@

In the multi-leader model, using a count of client connections to determine if
transaction is complete is problematic. First, the aggregation of completion
messages are based on the local aggregation point reporting to the transaction
leader.  There is no ability to detect or recover from a failure of this IOD
process that would report to the transaction leader. The current assumption
that the IOD layer will be an MPI program is not workable until MPI can fully
support fault detection and recovery including removing and adding processes.
Second, since the counts represent a begin/end transaction pair, it suffers
from a lack of accountability for missing object (shard) writes. All it would
indicate is that a process started and ended a transaction connection. There is
no validation that a process has done everything required of it.

@@

\DDT has addressed theses issues in a couple of ways. First, the
sub-coordinators each have a list of processes from which they expect messages.
Should a message be missed, it is noticed and corrective action can be taken.
Second, \DDT has the concept of sub-transactions. The messaging requirements
are illustrated in Figure~\ref{fig:optimized-protocol}.  Sub-transactions
represent finer grained operations than the entire output, \DDT can manage
multiple writes per client by using a sub-transaction to represent the output
for any item to the file (container). Because of how the sub-transactions are
managed, the singleton sub-transactions, ones in which only a single process
participates, must be declared before the transaction begins so that its
existence can be broadcast as part of the begin transaction message. This
ensures there is global knowledge that the sub-transaction is expected.  That
way if the coordinator (transaction leader) fails, whichever process takes over
that role knows to expect a completion message for that sub-transaction or the
overall transaction cannot complete. Global sub-transactions can be defined at
any time since they are a global, synchronized operation broadcasting their
existence. While this additional layer does introduce messaging, the overhead
is quite small.

@@

At a broader level, there are some concerns that were partially clarified
through conversations with the team.  Consider a shared file system across an
HPC data center. The current design maintains the metadata in its own
container. Since copying data from the IOD layer to the DAOS layer requires an
explicit persist call, how and when synchronizing the metadata across the
layers and potentially across machines occurs is undefined. Delaying
synchronization until an explicit persist is called will reduce the update
frequency, but delays the data visibility. Ideally, the metadata object would
need to be automatically persisted every time a container transaction is
persisted to the DAOS layer.

@@

\section{Acknowledgements}
The authors would like to thank Gary Grider, Eric Barton, Quincey Koziol, and
John Bent for their early review comments and discussions that clarified the
details of the design, the intent of the design, and the future plans.

\includegraphics[scale=0.07]{logos/doe_logo}
\includegraphics[scale=0.30]{logos/snl_logo}
\includegraphics[scale=0.35]{logos/nnsa_logo}
Sandia National Laboratories is a multi-program laboratory managed and operated
by Sandia Corporation, a wholly owned subsidiary of Lockheed Martin
Corporation, for the U.S. Department of Energy's National Nuclear Security
Administration under contract DE-AC04-94AL85000.
